---
title: "Coefficients interpretations"
author: "Michael Gurkov"
date: "August 19, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r set_parameters, echo=FALSE}

setwd("C:\\Users\\Misha\\Documents\\SandBox")

```

\section{Intro}

Linear regression can include transformed (in a non linear way) variables. In this note we'll look at logarithmic transformation and analyze and interpret the coefficients.

\subsection{Linear model}

The linear model is  

\begin{equation}
Y = \alpha +\beta X
\end{equation}

In that case the interpretation of $\beta$ is if X changes by one unit then Y changes by $\beta$ units. We can see it with simple example:

\begin{equation*}
  Y(X+1) - Y(X) = [\alpha + \beta (X+1)] - [\alpha + \beta X] = \beta
\end{equation*}

<!-- The partial effect of on Y of incremental change in X will be:  -->
<!-- \begin{equation*} -->
<!--   \frac{\Delta Y}{\Delta X} = \beta \Rightarrow \beta = \frac{\Delta Y}{\Delta X} -->
<!-- \end{equation*} -->


\subsection{Linear-log model}
In this case a logarithmic transformation is applied to X, the model is

\begin{equation*}
  Y = \alpha +\beta ln(X)
\end{equation*}


In that case the interpretation of $\beta$ is if X changes by one percent then Y changes by $\frac{\beta}{100}$ units. The example:

\begin{equation*}
  Y(1.01X) - Y(X) = [\alpha + \beta ln(1.01X)] - [\alpha + \beta ln(X)] = \beta ln(1.01) \approx \frac{\beta}{100}
\end{equation*}


<!-- The partial effect of on Y of incremental change in X will be: -->
<!-- \begin{equation*} -->
<!--   \frac{\Delta Y}{\Delta X} = \frac{\beta}{X}\Rightarrow \beta = \frac{\Delta Y}{\Delta X} \cdot X -->
<!-- \end{equation*} -->