---
title: "Panel regression in R"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r Import_libraries}

library(plm)

library(tidyverse)

library(ggplot2)

library(cowplot)

library(glue)

```


```{r define_functions}

plot_panel_data = function(panel_df, x_var, y_var){
  
  time_plot = panel_df %>% 
  ggplot(aes(x = as.numeric(time), y = {{y_var}}, color = id)) + 
  geom_line() + 
  xlab(NULL) + ylab(NULL) + ggtitle("Over time")

id_plot = panel_df %>%
  ggplot(aes(x = id, y = {{y_var}}, color = time)) +
  geom_point() +
  xlab(NULL) + ylab(NULL) + ggtitle("Over id")

xy_plot = panel_df %>%
  ggplot(aes(x = {{x_var}}, y = {{y_var}})) +
  geom_point() +
  xlab(vars({{x_var}})) + ylab((vars({{y_var}})))

lower_panel = plot_grid(time_plot, id_plot)
  
panel_plot = plot_grid(xy_plot,lower_panel,nrow = 2) 

print(panel_plot)


}

```



Panel regression refers to a data with a cross section as well as time dimension. The example we'll work with is the monthly expenses (as function of income)of Alice,Bob and Eve. 



\section{Cross section fixed effects}

The model that we are looking at is:

$$ Y_{it} = \alpha_{i} + \beta x_{it}$$

Each state has a unique level of consumption and if we don't account for it we
get the wrong estimates. That is probably because 


```{r cross_section_case}

set.seed(123)

cross_section_case = list()

coef_list = list(
  alice = list(int = 100, mpc = 0.8),
  bob = list(int = 200, mpc = 0.8),
  eve  = list(int = 300, mpc = 0.8)
)

cross_section_case$reg_data = coef_list %>% 
  map_dfr(function(temp_ind){
    
    temp_data = tibble(income = ceiling(runif(5,500,900))) %>% 
      mutate(consumption = temp_ind$int + temp_ind$mpc * income) %>% 
      rownames_to_column(var = "time")
    
    return(temp_data)

  },.id = "id")

rm(coef_list)


```


```{r cross_section_case_plot_data, fig.height=8}

plot_panel_data(cross_section_case$reg_data,x_var = income, y_var = consumption)


```

```{r cross_section_case_estimation}

cross_section_case$formula = formula("consumption ~ income")

cross_section_case$lm = lm(formula = cross_section_case$formula,
                      data = cross_section_case$reg_data)

cross_section_case$pooling_reg = plm(
  formula = formula(cross_section_case$formula),
  data = cross_section_case$reg_data,
  index = c("id", "time"),
  model = "pooling"
)

cross_section_case$individual_reg = plm(
  formula = formula(cross_section_case$formula),
  data = cross_section_case$reg_data,
  index = c("id", "time"),
  model = "within",
  effect = "individual"
)


```

```{r cross_section_check_coeffs, echo=TRUE}

all.equal(coefficients(cross_section_case$pooling_reg)[2],
          coefficients(cross_section_case$individual_reg))

```

\subsection {Demeaning estimation}

One method of dealing with this issue is look at the deviation from the average consumption by state. That will take care of the
level difference. The "demeaning" should be performed on all the variables.


```{r cross_section_case_demean_estimation}

cross_section_case$demean_reg = cross_section_case$reg_data %>% 
  group_by(id) %>% 
  mutate(across(c(income, consumption), ~. - mean(.))) %>% 
  lm(formula = cross_section_case$formula, data = .)

cross_section_case$dummy_reg = cross_section_case$reg_data %>% 
  lm(formula = "consumption ~ income + id", data = .)



```

```{r cross_section_check_demean_coeffs, echo=TRUE}

all.equal(coefficients(cross_section_case$demean_reg)[2],
          coefficients(cross_section_case$dummy_reg)[2],
          coefficients(cross_section_case$individual_reg))

```

\section{Time fixed effects}

The model that we are looking at is:

$$ Y_{it} = \alpha +  \gamma_{t} + \beta x_{it}$$

Now each time period has a unique level of consumption and if we don't account for
it we get the wrong estimates.


```{r time_case}

set.seed(123)

time_case = list()

coef_list = list(
  alice = list(int = 200, mpc = 0.8),
  bob = list(int = 200, mpc = 0.8),
  eve  = list(int = 200, mpc = 0.8)
)

time_case$reg_data = coef_list %>% 
  map_dfr(function(temp_ind){
    
    temp_data = tibble(income = ceiling(runif(5,500,900))) %>% 
      mutate(consumption = temp_ind$int + temp_ind$mpc * income) %>% 
      rownames_to_column(var = "time")
    
    return(temp_data)

  },.id = "id") %>% 
  mutate(consumption = case_when(time == 1 ~ consumption + 100,
                                 time == 2 ~ consumption + 200,
                                 time == 3 ~ consumption - 100,
                                 time == 4 ~ consumption - 200,
                                 time == 5 ~ consumption + 300))

rm(coef_list)


```


```{r time_case_plot_data, fig.height=8}

plot_panel_data(time_case$reg_data,x_var = income, y_var = consumption)


```

```{r time_case_estimation}

time_case$formula = formula("consumption ~ income")

time_case$lm = lm(formula = time_case$formula,
                      data = time_case$reg_data)

time_case$pooling_reg = plm(
  formula = formula(time_case$formula),
  data = time_case$reg_data,
  index = c("id", "time"),
  model = "pooling"
)

time_case$time_reg = plm(
  formula = formula(time_case$formula),
  data = time_case$reg_data,
  index = c("id", "time"),
  model = "within",
  effect = "time"
)


```

```{r time_check_coeffs, echo=TRUE}

all.equal(coefficients(time_case$pooling_reg)[2],
          coefficients(time_case$time_reg))

```

\subsection {Demeaning estimation}

One method of dealing with this issue is look at the deviation from the average consumption by state. That will take care of the
level difference. The "demeaning" should be performed on all the variables.


```{r time_case_demean_estimation}

time_case$demean_reg = time_case$reg_data %>% 
  group_by(time) %>% 
  mutate(across(c(income, consumption), ~. - mean(.))) %>% 
  lm(formula = time_case$formula, data = .)

time_case$dummy_reg = time_case$reg_data %>% 
  lm(formula = "consumption ~ income + time", data = .)



```

```{r time_check_demean_coeffs, echo=TRUE}

all.equal(coefficients(time_case$demean_reg)[2],
          coefficients(time_case$dummy_reg)[2],
          coefficients(time_case$time_reg))

```

\section{Cross section and time effect}


```{r cross_section_time_case}

set.seed(123)

cross_section_time_case = list()

coef_list = list(
  alice = list(int = 100, mpc = 0.8),
  bob = list(int = 200, mpc = 0.8),
  eve  = list(int = 300, mpc = 0.8)
)

cross_section_time_case$reg_data = coef_list %>% 
  map_dfr(function(temp_ind){
    
    temp_data = tibble(income = ceiling(runif(5,500,900))) %>% 
      mutate(consumption = temp_ind$int + temp_ind$mpc * income) %>% 
      rownames_to_column(var = "time")
    
    return(temp_data)

  },.id = "id") %>% 
  mutate(consumption = case_when(time == 1 ~ consumption + 100,
                                 time == 2 ~ consumption + 200,
                                 time == 3 ~ consumption - 100,
                                 time == 4 ~ consumption - 200,
                                 time == 5 ~ consumption + 300))

rm(coef_list)


```


```{r cross_section_time_case_plot_data, fig.height=8}

plot_panel_data(cross_section_time_case$reg_data,
                x_var = income,
                y_var = consumption)


```

```{r cross_section_time_case_estimation}

cross_section_time_case$formula = formula("consumption ~ income")

cross_section_time_case$lm = lm(formula = cross_section_time_case$formula,
                      data = cross_section_time_case$reg_data)

cross_section_time_case$pooling_reg = plm(
  formula = formula(cross_section_time_case$formula),
  data = cross_section_time_case$reg_data,
  index = c("id", "time"),
  model = "pooling"
)

cross_section_time_case$fe_reg = plm(
  formula = formula(cross_section_time_case$formula),
  data = cross_section_time_case$reg_data,
  index = c("id", "time"),
  model = "within",
  effect = "twoways"
)


```

```{r cross_section_time_check_coeffs, echo=TRUE}

all.equal(coefficients(cross_section_time_case$pooling_reg)[2],
          coefficients(cross_section_time_case$fe_reg))

```


```{r cross_section_time_demean_estimation}

cross_section_time_case$demean_reg = cross_section_time_case$reg_data %>% 
  group_by(time) %>% 
  mutate(across(c(income, consumption), ~. - mean(.))) %>%
  group_by(id) %>% 
  mutate(across(c(income, consumption), ~. - mean(.))) %>% 
  lm(formula = cross_section_time_case$formula, data = .)

cross_section_time_case$dummy_reg = cross_section_time_case$reg_data %>% 
  lm(formula = "consumption ~ income + id + time", data = .)



```

```{r cross_section_time_check_demean_coeffs, echo=TRUE}

all.equal(coefficients(cross_section_time_case$demean_reg)[2],
          coefficients(cross_section_time_case$dummy_reg)[2],
          coefficients(cross_section_time_case$fe_reg))

```


\section {Specific plm issues}

```{r cross_section_conflict_case}

coef_list = list(
  alice = list(int = 100, mpc = 0.8),
  bob = list(int = 200, mpc = 0.8),
  eve  = list(int = 300, mpc = 0.8)
)

cross_fe_data = coef_list %>% 
  map_dfr(function(temp_ind){
    
    temp_data = tibble(income = ceiling(runif(5,500,900))) %>% 
      mutate(consumption = temp_ind$int + temp_ind$mpc * income) %>% 
      rownames_to_column(var = "time")
    
    return(temp_data)

  },.id = "id") %>% 
  mutate(id_fe = if_else(id == "bob",1,0)) %>% 
  mutate(time_fe = if_else(time == 3,1,0)) %>%
  mutate(consumption = consumption + 300 * id_fe) %>% 
  mutate(consumption = consumption - 0.2 * income * time_fe)

rm(coef_list)


```

```{r cross_section_conflict_estimation}

lm_reg = lm(
  formula = formula("consumption ~ income*id_fe + income*time_fe + id + time"),
  data = cross_fe_data
)

plm_reg = plm(
  formula = formula("consumption ~ income*id_fe + income*time_fe"),
  data = cross_fe_data,
  index = c("id", "time"),
  model = "within",
  effect = "twoways"
)


```

```{r cross_section_conflict_case_check_coeffs, echo=TRUE}

all.equal(coefficients(lm_reg)[c("income","income:id_fe","income:time_fe")],
          coefficients(plm_reg)[c("income","income:id_fe","income:time_fe")])

```




\section{Degenerate case}

In case where the income (independent var) is the same for all the individuals
we'll get an identity between income levels and time periods (in each period)
all the individuals have the same income. In that case pooling regression and individual effect regression will give the same coeffs.

```{r cross_section_degenerate_case}

cross_section_degenerate_case = list()

coef_list = list(
  alice = list(int = 100, mpc = 0.8),
  bob = list(int = 200, mpc = 0.8),
  eve  = list(int = 300, mpc = 0.8)
)

cross_section_degenerate_case$reg_data = coef_list %>% 
  map_dfr(function(temp_ind){
    
    temp_data = tibble(income = seq(from = 500, to = 900, by = 100)) %>% 
      mutate(consumption = temp_ind$int + temp_ind$mpc * income) %>% 
      rownames_to_column(var = "time")
    
    return(temp_data)

  },.id = "id")

rm(coef_list)


```


```{r cross_section_degenerate_case_plot_data}

plot_panel_data(cross_section_degenerate_case$reg_data,x_var = income, y_var = consumption)


```


```{r cross_section_degenerate_case_estimation}

cross_section_degenerate_case$formula = formula("consumption ~ income")

cross_section_degenerate_case$lm = lm(formula = cross_section_degenerate_case$formula,
                      data = cross_section_degenerate_case$reg_data)

cross_section_degenerate_case$pooling_reg = plm(
  formula = formula(cross_section_degenerate_case$formula),
  data = cross_section_degenerate_case$reg_data,
  index = c("id", "time"),
  model = "pooling"
)

cross_section_degenerate_case$individual_reg = plm(
  formula = formula(cross_section_degenerate_case$formula),
  data = cross_section_degenerate_case$reg_data,
  index = c("id", "time"),
  model = "within",
  effect = "individual"
)


```

```{r check_coeffs, echo=TRUE}

all.equal(coefficients(cross_section_degenerate_case$pooling_reg)[2],
          coefficients(cross_section_degenerate_case$individual_reg))

```


As for the time effects aspect the degenerate case will produce perfect linear
combination since we'll essentially get three (by number of id's) "identical" 
observations for each time period.