---
title: "Panel regression in R"
author: "Michael Gurkov"
date: "December 5, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```


```{r Import_libraries}

library(plm)

library(tidyr)

library(ggplot2)

```


Panel regression refers to a data with a spatial as well as time dimension. The example we'll work with is the monthly expenses (as function of income)of Alice, Bob and Eve. 


\section{Plain vanila case}

```{r plain_vanila}

coef = list(Alice = list(int = 100, mpc = 0.8),
            Bob = list(int = 200, mpc = 0.8),
            Eve  = list(int = 300, mpc = 0.8))

income = seq.int(from = 1000,to = 5000,by = 100)

x_data = matrix(rep(income,3), ncol = 3)

x_data = cbind(1:nrow(x_data), x_data)

colnames(x_data) = c("Time","Alice","Bob","Eve")

x_data = gather(as.data.frame(x_data),"Name","Income",-1)

y_data = sapply(c("Alice","Bob","Eve"),
                function(name){coef[[name]]$int + coef[[name]]$mpc * income +
                    rnorm(length(income), sd = 40)})

y_data = cbind(1:nrow(y_data), y_data)

colnames(y_data) = c("Time","Alice","Bob","Eve")

y_data = gather(as.data.frame(y_data),"Name","Consumption",-1)

reg_data = merge.data.frame(x_data,y_data, by = c("Time","Name"))

reg_data = reg_data[order(reg_data$Time),c("Name","Time","Income","Consumption")]

reg = plm(formula = Consumption ~ Income,data = reg_data,model = "pooling")

reg_fe = plm(formula = Consumption ~ Income,data = reg_data,model = "within")


```


```{r plot_data, eval=FALSE}

ggplot(as.data.frame(y_data), aes(x = Time, y = Consumption, color = Name)) + 
  geom_line() + 
  theme_bw()

```


In this case the marginal propensity to consume (slope of the population regression line) as well as autonomous consumption (intercept of the population regression line) will be the same 
